---
title       : Intro to R for the C# Developer
subtitle    :  (Use arrow keys to navigate)
author      : Elizabeth Groom
job         : Software Engineer at Cartegraph\Data enthusiast
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---


## About me

Elizabeth Groom
Software Engineer Cartegraph in Dubuque since 2003

email me: eagroom@yahoo.com  
find this presentation on github at: [https://github.com/eagroom/IntroToRPresentation] (https://github.com/eagroom/IntroToRPresentation)  

`presentation built using slidfy in R` 

--- 

## Why this is important

Consider the case study presented by Keith Baggerly [The Anil Potti Story] (http://videolectures.net/cancerbioinformatics2010_baggerly_irrh/)

"Intuition in high through put biology is poor" -Keith Baggerly 

Article: "Genomic signatures to guide the use of chemotherapeutics in the treat of Cancer."
  Auther: Potti et al (2006), 
  Magazine: Nature Medicine, 12:1294-1300

The idea is they developed a tool that can be used to predict weather a patient will respond to chemotherapies by knowing What drugs are most effective against which cells.  

<b>Amazing!</b>

---

## Why Cont.

This was very exciting to M.D. Anderson who asked Baggerly's group to evaluate the Potti group's tools for predicting whether given patients would respond to different chemotherapies

What Baggerly found was poor data analysis as several steps along the way.  Easy things like mis-labeling of data.  But this caused patients to not get the drugs that it is predicted their cancer would respond best to!

"Intuition in high through put biology is poor" in other words... we need impartial engineers who care deeply about the process and the math data is curated through so that the doctors can geneticist can learn factual things from the data.

---

## What is R

> * R is an <b>interactive</b> system for statistical computation and graphics. It consists of a language `plus a run-time environment with graphics`, a debugger, access to certain system functions, and the ability to run programs stored in script files.

> * R is a dialect of the S language.

> * S is a language that was developed by John Chambers and others at Bell Labs. It was initiated in 1976 as an internal statistical analysis environment-originally implemented as Fortran libraries. (In 1988 the system was rewritten in C)

> * "[W]e wanted users to be able to begin in an interactive environment, where they did not consciously think of themselves as programming. Then as their needs became clearer and their sophistication increased, they should be able to slide gradually into programming, when the language and system aspects would become more important." John Chambers



---

## What is R Cont.

* In 1991 Ross Ihaka and Robert Gentleman created R

* Runs on almost any standard computing platform/OS (even on the PlayStation 3)

* Frequent releases (annual + bugfix releases); active development.

---

## The Tools

* R comes with an IDE.  Its not bad.

* Recommend RStudio. It's <b>`free`</b> and includes a lot of great features like syntax highlight, auto completion and integrated help.

---

## Variable assignment

```{r}
x<- 1

```

The assignment operator is an "<-".  You can read it as "x get's 1

```{r}
1 -> x #is also valid

```

---

## Two more ways to Assign variables

1. "=".  Works in "most" contexts
```{r}
x = 1
```

2. assign function
```{r}
assign("x", 1)
```

This leads me to the first thing that i learned the hard way about R. `There are at least 3 ways to do any one thing in R and they work in "most" context.`

---

## Auto print

To print out our value of x:

```{r}
x<- 10
x
```

---

## vectors

Notice how X prints out : [1] 10

The simplest structure in R is the numeric vector. The number in brackets denotes the number of items in the vector.  

We can check the vector type with the function class

```{r}
x

class(x)
```

---

## vectors cont.

R has five basic or "atomic" classes of objects:
* character
* numeric (real numbers)
* integer
* complex
* logical (True/False)

A vector can only contain objects of the same class

---

## Data frames - The most useful object

A data frame for all intents and purposes in a table.
* Each columns is a vector of the same type, but each column can be different
* Each column has the same length
* Columns can have labels

Let's Check out one of the the sample Data sets in R: mtCars

---

## Data frames

Data Frames have some helpful functions to help you get to know your data:

The str function: displays the structure of an object
```{r}
str(mtcars)
```

---

## Data frames

The summary function: displays some summary information about the object
```{r}
summary(mtcars)
```

1st Qu. = the first quantile- means 25% observations are below this 
3rd Qu. = the third quantile - means 75% observations are below this
Median = The middle point. Half the numbers are above this value
Mean = average

---

#Data frames

Let's confirm the mean of the miles per gallon column (20.09)

to get the just the mpg column we use the $

```{r}
 mtcars$mpg
```

---

then we can sum the column and get its length

```{r}
sumMPG <- sum(mtcars$mpg)
lengthMPG <- length(mtcars$mpg)

sumMPG/lengthMPG

```

---

```{r}
round(sumMPG/lengthMPG, 2)
```

Perfect!  That matches the summary value of 20.09

---

## Data frames

Head and Tail function let you see the top, or bottom rows of data.  By default they show 6 rows.
```{r}
head(mtcars)
tail(mtcars, 2)
```

---

## A real Data set

Let's take a look at the survey results from the 2015 
[http://stackoverflow.com/research/developer-survey-2015](Survey result)

Steps:
1. set the working directory
  point out: escape back slashes with double slashes like in c# or use the /
2. read the data from the file
3. show the environment variables
4. show the quick text with tab
5. factors
7. packages
8. help
9. ...parameter
10. optional parameters

---

## Reusable scripts

Now lets say I am interested in what countries the  over 60 crowd are in.

Point out:
comments
in a rerun script use function require
method returns
looping
sourcing the script

---

## R is a n interactive langue

It seems that the reproducibility of

---

## summary

 "As you might be aware, R language is an open-source language for programmers, and the language is the product of collaborative evolution from a combination of brilliant minds too numerous to count. Maybe the best result of this is that R Language compilers have been designed keeping a newbie programmer in mind, which makes it an easy language to adapt. However, numerous minds bring numerous differences as well, and due to that, the language has been facing some critical issues in performance, which you will come to know more about from reading this tutorial."
 
 -[https://blog.udemy.com/r-tutorial/](Laran Joseph)

---

##Questions?

-----


install.packages("devtools")
library(devtools)
install_github("slidify", "ramnathv")
install_github("slidifyLibraries", "ramnathv")
library(slidify)

slidify("index.Rmd")
---



